Transfer learning for \'echelle spectroscopy with \texttt{blas\'e}

### 2022-02-23 paper/paper1/ms.tex
Add the astronomy unified thesaurus UAT keywords, probably too many
fill in the references and edit the introduction from yesterday

\keywords{High resolution spectroscopy (2096)}

@@

Tens of thousands or more individial spectral lines give rise to a sea of undulations that imbue each stellar spectrum with its characteristic appearance.  The identification and understanding of these lines has consumed a chief discipline of astrophysics over the last century.  The field grew from by-eye catalogs of stellar templates (cite \censor{XX} Cecilia Payne, Cannon) to modern pre-computed synthetic forward models including millions or billions of lines (cite \censor{XX}, Husser, Marley, HITRAN, HITEMP, Exomol, radis, Hargraves).  But as technology has improved, our data and models have become more voluminous, more precise, and more complicated.  The mere act of comparing models to data has become a computational challenge as much as a scientific one.  Here we introduce a new state-of-the-art framework \texttt{blas\'e} aimed at solving computational, statistical, and scientific challenges associated with data-model comparisons for modern astronomical spectroscopy.

The \texttt{blas\'e} framework aims to take in an observed astronomical spectrum and report back an extensible semi-empirical spectral forward model, replete with the postition, amplitude, width, and shape of all of its spectral lines, automatically.  Solving this problem is hard, for at least five reasons.  First, the spectral lines may overlap, and so the assignment of one line may be partially degenerate with the assignment of some other adjacent line.  Second, the lines exhibit a large dynamic range in their allowed extents, with extremely wide line wings such as Sodium or Hydrogen lines blending into the continuum, such that the placement of the continuum level may become ill-defined.  Third, the extent of line blending and realized line shape depends strongly on the spectral resolution of the spectrograph, the rotational broadening of the star, and possibly the instrumental configuration at the time of observation.  Fourth, telluric absorption lines commingle with the astronomical spectral lines of interest, censoring some spectral regions entirely, or partially confounding other lines with chance allignments.  And finally, our prediction for where lines should be located may be incomplete, and the decision of where to place lines becomes an uncertain choice that depends on the conditions and assumptions of the surrounding lines.

Addressing these challenges forms the backbone of spectral calibration, an increasingly valuable specialty as the deficits in our models become intolerable with greater data quantity and quality.  Luckily, many scientific applications in astrophysics do not need the technically demanding noise-free template, nor catalog-of-all-spectral-lines.  A few lines suffice.  For those applications, human inspection of isolated lines and semi-automated equivalent width determination has been and will remain adequate.

But many new and important questions in the fields of stars and exoplanets aspire to reach the margins of what the data can inform.  For example, exoplanet cross-correlation spectroscopy (cite \censor{xx} Jayne Birkby, etc.) hinges on accurate templates to detect and characterize the atmospheres of exoplanets.  Imperfections in these templates can mute the perceived signal strength of these atmospheric features.  In extreme-precision-RV (EPRV) applications, the joint modeling of microtellurics may require an accurate semi-empirical model of both the star and the Earth's atmospheric absorption at that fleeting moment in time.  In the case of Doppler imaging, an accurate underlying template is needed to detect longitudinally symmetric structures such as polar spots (cite \censor{XX} Vogt, Roettenbacher) or zonal bands (cite \censor{XX}, Crossfield, Apai).  The \texttt{blas\'e} framework is suitable for these and many other applications for which an accurate noise-free template is desired, or for which the properties of a large number of individual spectral lines is sought.

Existing frameworks have already overcome aspects of these challenges, or have been purpose-built for specific applications.  The \texttt{specmatch} template matching tool produces noise-free nearest neighbor templates given an input spectrum \citep{2015PhDT........82P}.  Several practical barriers limit the accuracy of using precomputed synthetic spectral models. First and foremost, real stars are usually more complicated than our simplified models of them. Real spectra often vary over more dimensions than specified. Conspicuous examples of these hidden variables can be found in protostars: starspots, accretion veiling, dust extinction, and magnetic Zeeman splitting. Jointly modeling all of these phenomena alongside the intrinsic stellar photosphere is challenging.

The empirical version, \texttt{specmatch-emp} \citep{2017ApJ...836...77Y} matched spectra better than the synthetic templates, but was still too rigid for some applications and required the assembly of hundreds of standardized high signal-to-noise-ratio templates, ideally with low intrinsic rotational broadening.  Such a large number of high-quality templates has not yet been established in the near infrared.

@@

The \texttt{starfish} framework \citep{czekala15} provides a robust likelihood function for data-model comparisons, and retires many of the problems in this domain.  \texttt{starfish} pioneered the use of whole-spectrum fitting with imperfect models.  In particular, it addressed the problem of what to do when the underlying atomic and molecular data was wrong or approximate or missing.  It has been extended to inferring starspot physical properties \citep{2017ApJ...836..200G}, measuring veiling in Class 0 protostars \citep{2018ApJ...862...85G}, and quantifying imperfections in brown dwarf models (cite \censor{ZJ BD paper}).  The Spectral Inference Crank, \texttt{sick}, shares similar aims as \texttt{starfish}, and provides additional useful grid search capabilities.

>>>>

\keywords{High resolution spectroscopy (2096), Stellar spectral lines (1630), Astronomy data modeling(1859), GPU Computing (1969), Calibration (2179), Radial Velocity (1332), Maximum likelihood estimation (1901), Deconvolution (1910), Atomic spectroscopy (2099), Stellar photospheres (1237)},

@@

Tens of thousands or more individual spectral lines give rise to a sea of undulations that imbue each stellar spectrum with its characteristic appearance.  The identification and understanding of these lines has defined a large category of astrophysics over the last century.  The field grew from by-eye catalogs of stellar templates \citep{1901AnHar..28..129C} to quantifying the role of atomic ionization balance \citep{1925PhDT.........1P}, to modern synthetic forward models including millions or billions of lines \citep{husser13, 2021ApJ...920...85M}.  As technology has improved, our data and models have become more voluminous, more precise, and more complicated.  The mere act of comparing models to observed spectra can resemble a computational statistics challenge as much as a scientific one.  Here we introduce a machine-learning-based framework \texttt{blas\'e} aimed at solving computational, statistical, and scientific challenges associated with data-model comparisons for modern astronomical spectroscopy.

The \texttt{blas\'e} framework aims to take in an observed astronomical spectrum and report back an extensible semi-empirical spectral forward model, replete with the postition, amplitude, width, and shape of all of its spectral lines, automatically.  Solving this problem is hard, for at least four reasons.  First, the spectral lines may overlap, and so the assignment of one line may be partially degenerate with the assignment of some other adjacent line.  Second, extremely wide line wings blend into the continuum, such that the placement of the continuum level may become ill-defined.  Third, the extent of line blending and realized line shape depends strongly on the spectral resolution of the spectrograph, the rotational broadening of the star, and possibly the instrumental configuration at the time of observation.  Finally, telluric absorption lines commingle with the astronomical spectral lines of interest, censoring some spectral regions entirely, or partially confounding other lines with chance allignments.

Addressing these and other challenges forms the backbone of spectral calibration, an increasingly valuable specialty as the deficits in our models become intolerable with greater data quantity and quality.  Luckily, many scientific applications in astrophysics do not need the technically demanding noise-free template, nor catalog-of-all-spectral-lines.  A few lines suffice.  For those applications, human inspection of isolated lines and semi-automated equivalent width determination has been and will remain adequate.

But many new and important questions in the fields of stars and exoplanets aspire to reach the margins of what the data can inform.  For example, exoplanet cross-correlation spectroscopy \citep{2013MNRAS.436L..35B} hinges on accurate molecular spectral templates to detect and characterize the atmospheres of exoplanets.  Imperfections in these templates can mute the perceived signal strength of these atmospheric features \citep{2015A&A...575A..20H}.  
In extreme precision radial velocity (EPRV) applications, addressing telluric absorption at the $\mathrm{cm/s}$ level may require joint modeling of the star and the Earth's atmospheric absorption before convolution with an instrumental kernel\footnote{\url{https://speakerdeck.com/dwhgg/telluric-line-hack-week-wrap-up}}.  
In the case of Doppler imaging, an accurate underlying template is needed to detect longitudinally symmetric structures such as polar spots \citep{1983PASP...95..565V, roettenbacher16} or zonal bands \citep{Crossfield14,2021ApJ...906...64A}.  The \texttt{blas\'e} framework is suitable for these and many other applications for which an accurate \emph{evaluable} noise-free template is desired, or for which the properties of a large number of individual spectral lines is sought.

Existing frameworks have already overcome aspects of these challenges, or have been purpose-built for specific applications.  The \texttt{specmatch} synthetic template matching tool produces noise-free nearest neighbor templates given an input spectrum \citep{2015PhDT........82P}.  Several practical barriers limit the accuracy of using precomputed synthetic spectral models alone. First and foremost, real stars are usually more complicated than our simplified models of them. Real spectra often vary over more dimensions that our models do.  Conspicuous examples of these hidden variables can be found in protostars: starspots, accretion veiling, dust extinction, and magnetic Zeeman splitting. Jointly modeling all of these phenomena alongside the intrinsic stellar photosphere is challenging.

The empirical version, \texttt{specmatch-emp} \citep{2017ApJ...836...77Y} matched spectra better than the synthetic templates, but is still too rigid for some applications and requires the assembly of hundreds of standardized high signal-to-noise-ratio templates, ideally with low intrinsic rotational broadening.  Such a large number of high-quality templates has not yet been established in the near infrared, and may not be feasible for brown dwarfs.

@@

The \texttt{starfish} framework \citep{czekala15} provides a robust likelihood function for data-model comparisons, and retires many of the problems in this domain.  \texttt{starfish} pioneered the use of whole-spectrum fitting with resilience to model imperfections by addressing the problem of what to do when the underlying atomic and molecular data was wrong or approximate or missing.  It has been extended to inferring starspot physical properties \citep{2017ApJ...836..200G}, measuring veiling in Class 0 protostars \citep{2018ApJ...862...85G}, and quantifying imperfections in brown dwarf models \citep{2021ApJ...921...95Z}.  The Spectral Inference Crank \citep[\texttt{sick},][]{2016ApJS..223....8C} shares similar aims as \texttt{starfish}, and provides additional useful grid search capabilities.
